Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-18 ind_files/05_unification_discussion.tex
Minor phrasing changes

If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes obtained by `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes holds, then the exact same steps carried out in Appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} (with sums replaced by integrals and derivatives replaced by functional derivatives as appropriate) as the `fundamental theorems' for the evolution of quantitative traits. Indeed,~\cite{week_white_2021} have recently proposed exactly the It\^o formula we would need via a heuristic It\^o multiplication table for $L^2(\mathbb{R}, m)$ valued processes with a.s. finite Hilbert-Schmidt norm and have shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by~\cite{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like abstract gibberish to you, don't worry too much about the details - The essence is that~\cite{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility over mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied.}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to certain SPDEs,~\cite{week_white_2021} have arrived at precisely the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} obtained by `taking the limit' $m \to \infty$ in \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} from the `SPDE side' under certain additional assumptions required to ensure existence/uniqueness of solutions to the relevant SPDEs. Equation \eqref{disc_functional_field_eqns} is the (functional) Fokker-Planck view of the stochastic processes for the evolutionary ecology of quantitative traits recently studied in~\cite{week_white_2021}. Like~\cite{week_white_2021}, I believe that rigorously establishing the precise criteria needed for general existence and uniqueness of solutions to equations of the form \eqref{disc_functional_field_eqns} (or the SPDE versions presented in~\cite{week_white_2021}) presents an important mathematical direction for future work that may yield new insights into both the biology of quantitative traits and the mathematics of measure-valued stochastic processes.

More generally, the well-known intimate relation between It\^o SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) is extremely useful, and indeed has been extensively exploited~\citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009}, because many problems are much more easily attacked using probabilistic tools that operate on It\^o process, whereas others are more amenable to study using tools from PDE theory that operate on the Fokker-Planck equations. In the infinite-dimensional case, working with the field theory through functional Fokker-Planck equations such as \eqref{disc_functional_field_eqns} allows us to use tools with an essentially `differential equation' flavor such as spectral methods (\cite{rogers_demographic_2012, rogers_modes_2015}; See also Appendix \ref{App_examples} for a general model-independent pedagogical example for studying phenotypic clustering and evolutionary branching), slow manifold (`adiabatic') approximations~\citep{parsons_dimension_2017}, and WKB style expansions~(\cite{rogers_demographic_2012,assaf_wkb_2017}; though it bears noting that WKB theory is closely related to large deviation principles that occur in the study of Markov processes and come in a distinctly `probabilistic' flavor, see~\cite{dembo_large_1998}). On the other hand, many biologically interesting questions are much easier to approach from the branching processes/SPDE side using tools that have an essentially `probabilistic' flavor such as duality~\citep{dawson_stochastic_1975,greenman_duality_2020}, infinitesimal generators~\citep{ethier_markov_1986,etheridge_mathematical_2011}, martingale theory~\citep{champagnat_unifying_2006,etheridge_mathematical_2011}, and weak/mild solutions to SPDEs~\citep{week_white_2021}. Rigorously establishing and exploiting relations between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and SPDEs of the form \eqref{functional_langevin} (or those studied in~\cite{week_white_2021}) for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between complementary views of the same object to suit the problem at hand\footnote{Indeed, the physics literature has long recognized this. Physicists regularly exploit the connections between stochastic/quantum field equations and SPDEs, often studying SPDEs in \emph{classical} systems using tools and language from quantum field theory~\citep{hochberg_effective_1999,baez_quantum_2018}.}.

>>>>

If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes obtained by `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes holds, then the exact same steps carried out in Appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} (with sums replaced by integrals and derivatives replaced by functional derivatives as appropriate) as the `fundamental theorems' for the evolution of quantitative traits. Indeed,~\cite{week_white_2021} have recently proposed exactly the It\^o formula we would need via a heuristic It\^o multiplication table for $L^2(\mathbb{R}, m)$ valued processes with a.s. finite Hilbert-Schmidt norm and have shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by~\cite{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like abstract gibberish to you, don't worry too much about the details - The essence is that~\cite{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility and actually formulating the problem in formal terms rather than on mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied. The required properties will likely be fulfilled only for some `nice' subspace of $\mathcal{M}_K(\mathcal{T})$ and not the entire space, and future, more rigorous work should focus on figuring out which functions are included in this subspace.}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to an SPDE describing populations with Gaussian trait distributions,~\cite{week_white_2021} have arrived at precisely the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} obtained by `taking the limit' $m \to \infty$ in \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} for the special case in which the type-level quantity $f$ is simply the value of the quantitative trait under study. Equation \eqref{disc_functional_field_eqns} is the `(functional) Fokker-Planck view' of the stochastic processes for the evolutionary ecology of quantitative traits that~\cite{week_white_2021} study from the `SPDE view', and indeed the two approaches are exactly complementary. Like~\cite{week_white_2021}, I believe that rigorously establishing the precise criteria needed for general existence and uniqueness of solutions to equations of the form \eqref{disc_functional_field_eqns} (or the SPDE versions presented in~\cite{week_white_2021}) presents an important mathematical direction for future work that may yield new insights into both the biology of quantitative traits and the mathematics of measure-valued stochastic processes.

More generally, the well-known intimate relation between It\^o SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) is extremely useful, and indeed has been extensively exploited~\citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009}, because some problems are much more easily attacked using probabilistic tools that operate on It\^o processes, whereas others are more amenable to study using tools from PDE theory that operate on the Fokker-Planck equations. In the infinite-dimensional case, working with the field theory through functional Fokker-Planck equations such as \eqref{disc_functional_field_eqns} allows us to use tools with an essentially `differential equation' flavor such as spectral methods (\cite{rogers_demographic_2012, rogers_modes_2015}; See also Appendix \ref{App_examples} for a general model-independent pedagogical example for studying phenotypic clustering and evolutionary branching), slow manifold (`adiabatic') approximations~\citep{parsons_dimension_2017}, and WKB style expansions~(\cite{rogers_demographic_2012,assaf_wkb_2017}; though it bears noting that WKB theory is closely related to large deviation principles that occur in the study of Markov processes and come in a distinctly `probabilistic' flavor, see~\cite{dembo_large_1998}). On the other hand, many biologically interesting questions are much easier to approach from the branching processes/SPDE side using tools that have an essentially `probabilistic' flavor such as duality~\citep{dawson_stochastic_1975,greenman_duality_2020}, infinitesimal generators~\citep{ethier_markov_1986,etheridge_mathematical_2011}, martingale theory~\citep{champagnat_unifying_2006,etheridge_mathematical_2011}, and weak/mild solutions to SPDEs~\citep{week_white_2021}. Rigorously establishing and systematically exploiting relations between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and the sort of general SPDEs studied in~\cite{week_white_2021} for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between complementary views of the same object to suit the problem at hand\footnote{The physics literature has long recognized this. Physicists regularly exploit the connections between stochastic/quantum field equations and SPDEs, often studying SPDEs in \emph{classical} systems using tools and language from quantum field theory~\citep{hochberg_effective_1999,baez_quantum_2018}.}.

### 2023-03-18 ind_files/06_nontechnical_discussion.tex
Minor phrasing changes

In Chapter \ref{chap_infD_processes}, I also postulate a `stochastic field theory' approach to modelling the evolution of quantitative traits in finite, fluctuating populations. I then show that this approach is consistent with known frameworks in theoretical population biology, reproducing results from quantitative genetics in the infinite population limit. This formulation also highlights a (somewhat heuristic) approach to the study of spacetime stochastic processes and related mathematical objects that avoids measure-theoretic tools and may be of independent interest to applied mathematicians. Broadly similar stochastic field theoretic approaches have also been proposed in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014} and models of collective motion~\citep{o_laighleis_minimal_2018}. I  conjecture that the stochastic `fundamental equations' I derive in section \ref{sec_fun_theorems} for discrete traits should also hold for quantitative traits under mild assumptions on the trait space of the quantitative traits, though no proof is attempted in this thesis. We can, however, take encouragement from the observation that~\cite{week_white_2021} have already derived some special cases of these equations\footnote{In particular, equations (21b) and (21c) in \cite{week_white_2021} are precisely the $m \to \infty$ limit of my equations for changes in the mean value of a type-level quantity (equation \ref{nD_stochastic_Price}) and changes in the variance of a type-level quantity (equation \ref{nD_stochastic_Price_variance}) respectively for the special case in which the type-level quantity is the value of the quantitative trait being studied. See section \ref{sec_disc_field_eqns} for a more detailed technical discussion.} for quantitative traits using an approach grounded in the theory of measure-valued branching processes under certain technical assumptions on the stochastic process under study, the most biologically important assumption being that the traits are normally distributed in the population. My formalism and \cite{week_white_2021}'s formalism are complementary to each other, and reflect a deep duality between related mathematical structures that is well-appreciated in the applied mathematics literature (see section \ref{sec_disc_field_eqns}). These two complementary approaches to modeling stochastic processes have long been recognized in population genetics, where the (simpler analogs of the) general approach in models of fixed total population size go by the names `branching processes approach'~(\cite{week_white_2021}'s formalism; In the classic literature, seen in work like~\cite{haldane_mathematical_1927,fisher_distribution_1931}) and `diffusion theory approach'~(My formalism; In the classic literature, seen in work like~\cite{fisher_dominance_1923,wright_evolution_1931,kimura_problems_1957}).

@@

~\cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in my framework - Equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula, in principle for any moment of the distribution of the quantity $f$ in the population (though this quickly becomes too tedious to actually carry out in practice). If we additionally assume that the quantity $f$ follows a Gaussian distribution in the population, then the mean and variance completely characterize the distribution of $f$, and equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} together specify the complete stochastic dynamics of the system. Note that even in this case, actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated.

Previous studies indicate that in high dimensions, evolutionary birth-death models can exhibit a dizzying array of complicated phenomena, including limit cycles and evolutionary chaos~\citep{doebeli_diversity_2017}. I have also neglected any potential complications introduced by genotype-phenotype maps and genetic processes such as dominance, epistasis, and pleiotropy since they can (in principle) be incorporated while defining the per-capita birth and death rates and the equations I derive hold for arbitrarily complicated birth and death rates as long as they are of the functional form \eqref{nD_functional_forms_for_replicator}. However, while these facts may present impediments to formulating exact solutions to the equations I formulate, the point of these equations is not necessarily to solve them to begin with.

Indeed, the most important takeaway from this thesis is not in the solutions to the equations of section \ref{sec_fun_theorems} but in the formulation and the equations themselves. The equations I derive are very general, since part \ref{part_theory} makes essentially no assumptions other than density dependence, the impossibility of infinite growth starting from finite population size, and the ability to define per-capita birth and death rates, and thus, these birth and death rate function(al)s can in general be almost arbitrarily complicated. They thus apply to a very wide array of biological equations. Further, as we saw in section \ref{sec_fun_theorems}, the terms of these equations lend themselves to simple biological interpretation and make some general \emph{qualitative} predictions about how evolution should operate in finite, fluctuating populations. Namely, my equations show that evolution should operate similarly to how it operates in infinite populations, with the addition of some non-directional fluctuations (this is just drift) as well as an extra \emph{directional} force (noise-induced selection). Like the classical Price equation, the utility of the equations of section \ref{sec_fun_theorems} thus lies not (necessarily) in their solutions for specific models, but instead in their generality and the fact that their terms help us clearly think about the various evolutionary phenomena at play in biological populations~\citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general spirit of this thesis is thus in line with the general idea of trying to formulate `model-independent' eco-evolutionary theory that has recently been (rapidly) gaining popularity in the literature~\citep{grafen_formal_2014, queller_fundamental_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022,lion_extending_2023,allen_natural_2023}. However, there are some biologically important factors that I have neglected in my formalism. My framework works with unstructured populations, neglecting any potential effects of groups, age, class, sex, developmental stage, or space\footnote{Of course,  since position in continuous space is just a special case of a quantitative trait, we have technically incorporated space in a very limited sense: The formalism of chapter \ref{chap_infD_processes} can equally well describe a finite population of exactly identical individuals moving through one-dimensional space, and so can technically describe phenomena like range expansion of clonal populations. However, I am only noting this as a technicality - Such populations are likely somewhat `evolutionarily boring', since all individuals in the model must always be exactly identical in all aspects other than spatial location.}, all of which can lead to very complex and often surprising dynamics. Explicitly incorporating features such as population structure, sex, and space from first principles in an analytically tractable framework is a formidable task that may need innovative new mathematical and biological arguments, and presents a fantastic opportunity for future studies.

>>>>

In Chapter \ref{chap_infD_processes}, I also postulate a `stochastic field theory' approach to modelling the evolution of quantitative traits in finite, fluctuating populations. I then show that this approach is consistent with known frameworks in theoretical population biology, reproducing results from quantitative genetics in the infinite population limit. This formulation also highlights a (somewhat heuristic) approach to the study of spacetime stochastic processes and related mathematical objects that avoids measure-theoretic tools and may be of independent interest to applied mathematicians. Broadly similar stochastic field theoretic approaches have also been proposed in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014} and models of collective motion~\citep{o_laighleis_minimal_2018}. I  conjecture that the stochastic `fundamental equations' I derive in section \ref{sec_fun_theorems} for discrete traits should also hold for quantitative traits under mild assumptions on the trait space of the quantitative traits, though no proof is attempted in this thesis. We can, however, take encouragement from the observation that~\cite{week_white_2021} have already derived some special cases of these equations\footnote{In particular, equations (21b) and (21c) in \cite{week_white_2021} are precisely the $m \to \infty$ limit of my equations for changes in the mean value of a type-level quantity (equation \ref{nD_stochastic_Price}) and changes in the variance of a type-level quantity (equation \ref{nD_stochastic_Price_variance}) respectively for the special case in which the type-level quantity is the value of the quantitative trait being studied. See section \ref{sec_disc_field_eqns} for a more detailed technical discussion.} for quantitative traits using an approach grounded in the theory of measure-valued branching processes under certain technical assumptions on the stochastic process under study, the most biologically important assumption being that the traits are normally distributed in the population. My formalism and \cite{week_white_2021}'s formalism are complementary to each other, and reflect a deep duality between related mathematical structures that is well-appreciated in the applied mathematics literature (see section \ref{sec_disc_field_eqns}). These two complementary approaches to modeling stochastic processes have long been recognized in population genetics, where siminlar analogs of the broad idea of the approach for models of fixed total population size go by the names `branching processes approach'~(\cite{week_white_2021}'s formalism; In the classic literature, seen in work like~\cite{haldane_mathematical_1927,fisher_distribution_1931}) and `diffusion theory approach'~(My formalism; In the classic literature, seen in work like~\cite{wright_evolution_1931,kimura_problems_1957}).

@@

~\cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in my framework - Equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula, in principle for any moment of the distribution of the quantity $f$ in the population (though this quickly becomes too tedious to actually carry out in practice). If we additionally assume that the quantity $f$ follows a Gaussian distribution in the population, then the mean and variance completely characterize the distribution of $f$, and equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} together specify the complete stochastic dynamics of the system. Note that even in this case, actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that in high dimensions, evolutionary birth-death models can exhibit a dizzying array of complicated phenomena, including limit cycles and evolutionary chaos~\citep{doebeli_diversity_2017}. However, while this may present impediments to formulating exact solutions to the equations I formulate, the point of these equations is not necessarily to solve them to begin with.

Indeed, the most important takeaway from this thesis is not in the solutions to the equations of section \ref{sec_fun_theorems} but in the formulation and the equations themselves. The equations I derive are very general, since part \ref{part_theory} makes essentially no assumptions other than density dependence, the impossibility of infinite growth starting from finite population size, and the ability to define per-capita birth and death rates, and thus, these birth and death rate function(al)s can in general be almost arbitrarily complicated. They thus apply to a very wide array of biological populations. Further, as we saw in section \ref{sec_fun_theorems}, the terms of these equations lend themselves to simple biological interpretation and make some general \emph{qualitative} predictions about how evolution should operate in finite, fluctuating populations. Namely, my equations show that evolution should operate similarly to how it operates in infinite populations, with the addition of some non-directional fluctuations (this is just drift) as well as an extra \emph{directional} force (noise-induced selection). Like the classical Price equation, the utility of the equations of section \ref{sec_fun_theorems} thus lies not (necessarily) in their solutions for specific models, but instead in their generality and the fact that their terms help us clearly think about the various evolutionary phenomena at play in biological populations~\citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general spirit of this thesis is thus in line with the general idea of trying to formulate `model-independent' eco-evolutionary theory that has recently been (rapidly) gaining popularity in the literature~\citep{grafen_formal_2014, queller_fundamental_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022,lion_extending_2023,allen_natural_2023}.

There are many biologically important factors that I have neglected in my formalism as a first order approximation. My framework works with unstructured populations, neglecting any potential effects of groups, age, class, sex, developmental stage, or space\footnote{Of course,  since position in continuous space is just a special case of a quantitative trait, we have technically incorporated space in a very limited sense: The formalism of chapter \ref{chap_infD_processes} can equally well describe a finite population of exactly identical individuals moving through one-dimensional space, and so can technically describe phenomena like range expansion of clonal populations in one dimension. However, I am only noting this as a technicality - Such populations are likely somewhat `evolutionarily boring', since all individuals in the model must always be exactly identical in all aspects other than spatial location.}, all of which can lead to very complex and often surprising dynamics. Explicitly incorporating features such as population structure, sex, and space from first principles in an analytically tractable framework is a formidable task that may need innovative new mathematical and biological arguments, and presents a fantastic opportunity for future studies. I have also neglected any potential complications introduced by genotype-phenotype maps and genetic processes such as dominance, epistasis, and pleiotropy since they can (in principle) be incorporated into the definition of the per-capita birth and death rates in \eqref{nD_functional_forms_for_replicator}.
